---
title: "wood_dynamism"
author: 'Noël Clark'
date: "2026-06-22"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      out.width = "100%")
```

```{r load libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(plotly)
library(patchwork)
library(GGally)
library(lme4)
library(DHARMa)
library(lmerTest)
library(performance)
```

```{r load data, message=FALSE}
##### change directory if needed ######
data_dir <- "C:/Users/nc274936/OneDrive - The University of Montana/wood_dynamism/data"

dynamism_data <- read_csv(file.path(data_dir, "/Metrics/site_metrics.csv"))

dynamism_data <- dynamism_data[1:13,] %>% # for some reason we have duplicate rows in the csv
  bind_rows( # also want to add back the mod rock_WA site
    dynamism_data %>% slice(6) %>% 
      mutate(across(
        c(site_reach, site),
        ~ str_replace_all(.x, "_1", "_2")
        ))
    )

data_byyr <- read_csv(file.path(data_dir, "/Metrics/SelectMetricsByYear_06152026.csv"))
data_byyr <- data_byyr %>% 
  mutate(site_reach = paste0(site,reach))

wood_metrics <- data_byyr %>%
  mutate(wood_pieces_per_100m = (wood_pieces*100)/(channel_length),
         wood_accum_per_area = accum_area/channel_area) %>%
  group_by(site_reach) %>%
  summarize(avg_wood_pieces_per100m = mean(wood_pieces_per_100m),
            mean_wood_accum_per_area = mean(wood_accum_per_area))

all_data_byyr <- read.csv(file.path(data_dir, "/Metrics/ALL_metrics_06152026.csv"))
all_data_byyr <- all_data_byyr %>% 
  mutate(site_reach = paste0(site,reach))
```

# Simple Plots

## plot settings
```{r plot settings}
wood_colors = c("LW_Channel" = "#99C5B5", "no_LW_Channel" = "#BEA57D")
```

## wood values by site
```{r simple wood site plots}
##### wood values by site by year #####
data_byyr %>% 
  ggplot(aes(x = year, y = accum_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors) +
  theme(legend.position = "none") 

data_byyr %>% 
  ggplot(aes(x = year, y = wood_pieces, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

##### mean wood values by site #####
dynamism_data %>% 
  ggplot(aes(x = site, y = avg_wood_pieces_per100m, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors) +
  theme(legend.position = "none")

dynamism_data %>% 
  ggplot(aes(x = site, y = mean_wood_accum_per_area, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors)

```

## channel values by site
```{r simple channel site plots}
##### channel characteristics v time by site #####
data_byyr %>% 
  ggplot(aes(x = year, y = channel_length, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = channel_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) + 
  scale_color_manual(values = wood_colors)
```

## response values by site
```{r simple response values by site}
##### response metrics v time by site #####
data_byyr %>% 
  ggplot(aes(x = year, y = sinuosity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = complexity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = sdi, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

all_data_byyr %>% 
  ggplot(aes(x = year_to, y = net_change, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

all_data_byyr %>% 
  ggplot(aes(x = year_to, y = erosion_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

all_data_byyr %>% 
  ggplot(aes(x = year_to, y = deposition_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)
```

## dynamism values by site
```{r simple dynamism plots}
##### Sinuosity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Sinuosity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Sinuosity Dynamism vs. Avg Wood Metrics")
##### end 

##### Complexity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Complexity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Complexity Dynamism vs. Avg Wood Metrics")
##### end

##### SDI vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "SDI Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "SDI Dynamism vs. Avg Wood Metrics")
##### end

##### Net change vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Net Change Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Net Change Dynamism vs. Avg Wood Metrics")
##### end
```

## correlation plots and test 
### ggpairs uses the pearson correlation test by default
```{r correlation plot test}
# ggpairs uses pearson cor test by default 
data_byyr %>%
  mutate(wood_pieces_per_100m = (wood_pieces*100)/(channel_length),
         wood_accum_per_area = accum_area/channel_area) %>% 
  select(sinuosity, complexity, sdi, channel_length, channel_area, wood_pieces_per_100m, wood_accum_per_area) %>% 
  ggpairs(progress = F)

dynamism_data %>% 
  select(sinuosity_dynamism, complexity_dynamism, sdi_dynamism, net_change_dynamism, avg_wood_pieces_per100m, mean_wood_accum_per_area) %>% 
  ggpairs(progress = F)
```

## test null model for sinuosity 
```{r models}
mod_data <- dynamism_data %>% 
  filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
  mutate(site = str_replace_all(site, "_[:digit:]$", ""))

dyn_null <- lmer(sinuosity_dynamism ~ (1|site), mod_data)

plot(simulateResiduals(dyn_null))

summary(dyn_null)
coef(dyn_null)
r2(dyn_null)

site_intercepts <- coef(dyn_null)$site$"(Intercept)"

mod_data %>% 
  ggplot(aes(x=site, y=sinuosity_dynamism)) +
  geom_point() +
  geom_hline(yintercept = 0.010176) +
  geom_hline(yintercept = site_intercepts, linetype = "dashed", color = "gray32") +
  labs(title = "Null Model: Random Effect for Site",
       caption = "Black line is global mean, gray dashed lines are site-intercepts. Site explains ~21% of variation in sinuosity dynamism.")

```

## test timeseries trend model for channel area at Skunk Cr and Indian Cr
## !don't interpret
## !skunk model did not converge 
## !indian model terminated 
```{r timeseries trend model}
##### for skunk site #####
skunk_byyr <- data_byyr %>% 
  filter(site == "Skunk") 
# channel area over time
# model didn't converge fully?
skunk.ch.area.lm <- lm(channel_area ~ year + reach + (year*reach), skunk_byyr)

plot(simulateResiduals(skunk.ch.area.lm))

summary(skunk.ch.area.lm)
coef(skunk.ch.area.lm)
r2(skunk.ch.area.lm)

skunk_byyr %>% 
  ggplot(aes(year, channel_area, color = reach)) +
  geom_point() +
  geom_line(aes(group=reach)) +
  geom_smooth(method = "lm", formula = y ~ x + x + (x*x))

##### for indian site #####
indian_byyr <- data_byyr %>% 
  filter(site == "Indian") 
# channel area over time
# model fitting terminated with error!
indian.ch.area.lm <- lm(channel_area ~ year + reach + (year*reach), indian_byyr)

plot(simulateResiduals(indian.ch.area.lm))

summary(indian.ch.area.lm)
coef(indian.ch.area.lm)
r2(indian.ch.area.lm)

indian_byyr %>% 
  ggplot(aes(year, channel_area, color = reach)) +
  geom_point() +
  geom_line(aes(group=reach)) +
  geom_smooth(method = "lm", formula = y ~ x + x + (x*x))
```

