---
title: "wood_dynamism"
author: 'Noël Clark'
date: "2026-06-22"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      out.width = "100%")
```

```{r load libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(plotly)
library(patchwork)
library(GGally)
library(lme4)
library(DHARMa)
library(lmerTest)
library(performance)
library(MASS)
library(see)
library(AICcmodavg)
library(ggeffects)
library(sjPlot)
```

```{r load data, message=FALSE}
## change directory if needed ##
data_dir <- "C:/Users/nc274936/OneDrive - The University of Montana/wood_dynamism/data"
setwd("~/Library/CloudStorage/OneDrive-TheUniversityofMontana/Wood+/Wood+") # Kelley's working directory

dynamism_data <- read_csv(file.path(data_dir, "Metrics/site_metrics_07022026.csv"))
dynamism_data <- read_csv("site_metrics_07022026.csv") # Kelley's path

dynamism_data <- dynamism_data[1:13,] %>% # for some reason we have duplicate rows in the csv
  bind_rows( # also want to add back the mod rock_WA site
    dynamism_data %>% slice(6) %>% 
      mutate(across(
        c(site_reach, site),
        ~ str_replace_all(.x, "_1", "_2")
        ))
    ) 

data_byyr <- read_csv(file.path(data_dir, "/Metrics/SelectMetricsByYear_07022026.csv"))
data_byyr <- read_csv("SelectMetricsByYear_07022026.csv") # Kelley's path
data_byyr <- data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach)) %>% 
  mutate(ratio_accum_to_ch_area = accum_area/channel_area,
         piece_per_100m = (wood_pieces/channel_length)*100,
         site = factor(site),
         reach = factor(reach))


all_data_byyr <- read.csv(file.path(data_dir, "/Metrics/ALL_metrics_07022026.csv"))
all_data_byyr <- read.csv("ALL_metrics_07022026.csv") # Kelley's path
all_data_byyr <- all_data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach))
```

# Simple Plots
 
plot settings
```{r plot settings}
wood_colors = c("LW_Channel" = "#99C5B5", "no_LW_Channel" = "#BEA57D")
```

# visualizing distribution of response metrics 

support for the response values is positive & continuous
none of the response values are normally distributed - may require transformation?

support for the predictor is between 0 and 1 (proportion)
the predictor follows a negative binomial (?) distribution

```{r simple response value plots}
data_byyr %>% 
  ggplot(aes(x=sinuosity)) +
  geom_density() 

data_byyr %>% 
  ggplot(aes(x=complexity)) +
  geom_density()

data_byyr %>% 
  ggplot(aes(x=sdi)) +
  geom_density() 

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area)) +
  geom_density() 

data_byyr %>%  
  ggplot(aes(x=piece_per_100m)) +
  geom_density() 
```

## stage 1: wood as a response to site, reach, & time
```{r wood v time plots}
# how does standardized wood (the ratio of wood accumulation to channel area) change over time?
data_byyr %>%
  ggplot(aes(x=year, y=ratio_accum_to_ch_area, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach))

# how does the difference between standardized wood change within a site (between reaches) over time? (diff = LW - noLW)
data_byyr %>%
  group_by(site, year) %>% 
  arrange_all(by_group = "reach") %>% 
  mutate(accum_ratio_diff = lag(ratio_accum_to_ch_area, 1) - ratio_accum_to_ch_area) %>% 
  dplyr::select(site, year, accum_ratio_diff) %>% 
  drop_na(accum_ratio_diff) %>% 
  ggplot(aes(x=year, y=accum_ratio_diff, color = site)) +
  geom_point() +
  geom_line(aes(group = site))


# how does standardized wood pieces change over time?
data_byyr %>% 
  ggplot(aes(x=year, y=piece_per_100m, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach))

# how does the difference between standardized number of wood pieces change within a site (between reaches) over time? (diff = LW - noLW)
data_byyr %>%
  group_by(site, year) %>% 
  arrange_all(by_group = "reach") %>% 
  mutate(piece_100m_diff = lag(piece_per_100m, 1) - piece_per_100m) %>% 
  dplyr::select(site, year, piece_100m_diff) %>% 
  drop_na(piece_100m_diff) %>% 
  ggplot(aes(x=year, y=piece_100m_diff, color = site)) +
  geom_point() +
  geom_line(aes(group = site))
```
```{r simple wood site plots}
##### wood values by site by year #####
data_byyr %>% 
  ggplot(aes(x = year, y = ratio_accum_to_ch_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors) +
  theme(legend.position = "none") 

data_byyr %>% 
  ggplot(aes(x = year, y = piece_per_100m, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

##### mean wood values by site #####
dynamism_data %>% 
  ggplot(aes(x = site, y = avg_wood_pieces_per100m, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors) +
  theme(legend.position = "none")

dynamism_data %>% 
  ggplot(aes(x = site, y = mean_wood_accum_per_area, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors)

```


## stage 2: how do sinuosity, complexity, and sdi change as a function of standardized wood?

visualizing relationship between response v predictor (ratio_accum_to_ch_area)
within a site_reach combo the relationship is linear-ish? 
```{r visualize sin comp and sdi v wood}
data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sinuosity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=complexity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sdi, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))


## wood piece_per_100m
data_byyr %>% 
  ggplot(aes(x=piece_per_100m, y=sinuosity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))

data_byyr %>% 
  ggplot(aes(x=piece_per_100m, y=complexity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))

data_byyr %>% 
  ggplot(aes(x=piece_per_100m, y=sdi, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))
```

channel values by site
```{r simple channel site plots}
# channel characteristics v time by site
data_byyr %>% 
  ggplot(aes(x = year, y = channel_length, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = channel_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) + 
  scale_color_manual(values = wood_colors)
```

response values by site
```{r simple response values by site}
# response metrics v time by site
data_byyr %>% 
  ggplot(aes(x = year, y = sinuosity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = complexity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = sdi, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = net_change, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = erosion_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = deposition_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
```

dynamism values by site
```{r simple dynamism plots}
##### Sinuosity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Sinuosity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Sinuosity Dynamism vs. Avg Wood Metrics")
##### end 

##### Complexity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Complexity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Complexity Dynamism vs. Avg Wood Metrics")
##### end

##### SDI vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "SDI Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "SDI Dynamism vs. Avg Wood Metrics")
##### end

##### Net change vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Net Change Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Net Change Dynamism vs. Avg Wood Metrics")
##### end
```

```{r simple dynamism plots pt. 2}
library(grid)
library(rcartocolor)
library(dplyr)
library(patchwork)

## CLEANING UP
# Adjusting so all 3 metrics can be plotted on same graph
dynamism_cubed <- dynamism_data %>%
  pivot_longer(c(sinuosity_dynamism, sdi_dynamism, net_change_dynamism),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  mutate(Value_scaled = (scale(Value))) # calculated a separate z-score for each dynamism metric
  # will show dynamism values relative to each other

# Essentially, does wood accumulation change dynamism relative to it's normal range?

# Looking at colorblind friendly color schemes
rcartocolor::display_carto_all()
carto_pal(7, "Fall")

# Renaming labels
dynamism_cubed <- dynamism_cubed %>%
  mutate(
    Metric = recode(
      Metric,
      sinuosity_dynamism = "Sinuosity",
      sdi_dynamism = "SDI",
      net_change_dynamism = "Net Channel Change"
    ),
    reach = recode(
      reach,
      LW_Channel = "Large Wood",
      no_LW_Channel = "Control"
    ))
    
## PLOTS
# Good looking plots but lacking detail of site and reach
ggplot(dynamism_cubed, aes(avg_wood_pieces_per100m, Value_scaled, color = Metric)) +
  geom_line(size = 1)

ggplot(dynamism_cubed, aes(mean_wood_accum_per_area, Value_scaled, color = Metric)) +
  geom_line(size = 1)

# For avg_wood_pieces_per100m
avgwoodpieces <- ggplot(
  dynamism_cubed,
  aes(
    x = avg_wood_pieces_per100m,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
  facet_wrap(~reach) +
    facet_wrap(~reach, scales = "free_x") +
    labs(
    x = "Mean wood pieces (per 100m)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()


# For mean_wood_accum_per_area
meanwood <- ggplot(
  dynamism_cubed,
  aes(
    x = mean_wood_accum_per_area,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
    facet_wrap(~reach, scales = "free_x") +
    labs(
    x = "Mean wood accumulation (per area)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()

# Fixing x-axes so you can see them on their own scale since it's wacky in the above
meanwood_freex <- ggplot(
  dynamism_cubed,
  aes(
    x = mean_wood_accum_per_area,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(
    aes(group = Metric),
    method = "lm",
    se = FALSE,
    linewidth = 1.2
  ) +
  facet_wrap(~reach, scales = "free_x") +
    labs(
    x = "Mean wood accumulation (per area)",
    #y = "Standardized dynamism (z-score)", removing so it doesn't show up twice in patchwork
    color = "Dynamism Metric",
    shape = "Site"
  ) +
  theme_bw()+
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())


## FINAL PLOT
(meanwood + meanwood_freex) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Dynamism vs. Average Wood Metrics") &
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.x = unit(0.2, "cm"),
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0.1, "cm"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  ) &
  guides(
    color = guide_legend(
      title = "Dynamism Metric",
      ncol = 1
    ),
    shape = guide_legend(
      title = "Site",
      ncol = 1,
      override.aes = list(size = 3)
    )
  )
```


# correlation plots and test 

ggpairs uses the pearson correlation test by default
```{r correlation plot test}
# ggpairs uses pearson cor test by default 
data_byyr %>%
  dplyr::select(sinuosity, complexity, sdi, channel_length, channel_area, accum_area, ratio_accum_to_ch_area, wood_pieces, piece_per_100m) %>% 
  ggpairs(progress = F)

dynamism_data %>% 
  dplyr::select(sinuosity_dynamism, complexity_dynamism, sdi_dynamism, net_change_dynamism, avg_wood_pieces_per100m, mean_wood_accum_per_area) %>% 
  ggpairs(progress = F)
```

# Models

"random effects should reflect the design, but complexity should be justified by the amount of information available"
experimental design is such that reaches are nested within sites to account for multiple observations (across years) at reach_site combinations 

```{r create model df}
mod_data_byyr <- data_byyr %>% 
  mutate(ratio_accum_scaled = scale(ratio_accum_to_ch_area, 
                                    center = TRUE, # centers ratio values around mean ratio which is then set to 0 
                                    scale = TRUE), # scale changes a value to the # of sd from the mean value
         piece_100m_scaled = scale(piece_per_100m, 
                                   center = TRUE, 
                                   scale = TRUE),
         year_c = scale(year, # centers years around a mean (set to 0) value for year, makes for a more interpretable effect
                        center = TRUE, scale = FALSE)) %>% 
  filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
  mutate(site = str_replace_all(site, "_[:digit:]$", ""))
```

## modeling notes
- AICc is preferred for small sample sizes which is when the number of data points / number of parameters < 40 (n/k < 40). we have 126 observations, which for the null model means we're not in the small sample size world (k=3), but for all other models we are because we have to estimate at least 4 parameters. it's good to use the same criteria across models so we'll use AICc for all. the aictab function in the AICcmodavg does this for us.  
- all of our response variables are not normally distributed so we're using a generalized linear mixed model with a Gamma family and log link unless otherwise stated

## models for sinuosity
```{r specify sinuosity candidate model}
## null model: random effect of site only reflects hierarchical design 
# this will estimate a random intercept of sinuosity for every site (i.e. mean sinuosity per site)
null_sin <- glmer(sinuosity ~ (1|site), mod_data_byyr, family = Gamma(link = "log"))

## models with scaled ratio of wood accumulation
m1_sin <- glmer(sinuosity ~ ratio_accum_scaled + (1|site), mod_data_byyr, family = Gamma(link = "log"))
m2_sin <- glmer(sinuosity ~ piece_100m_scaled + (1|site), mod_data_byyr, family = Gamma(link = "log"))

m3_sin <- glmer(sinuosity ~ ratio_accum_scaled + piece_100m_scaled + (1|site), mod_data_byyr, family = Gamma(link = "log"))
m4_sin <- glmer(sinuosity ~ ratio_accum_scaled * piece_100m_scaled + (1|site), mod_data_byyr, family = Gamma(link = "log"))

## models with more complicated random effects structure
# these calculate both a random intercept and a random slope within a site
m5_sin <- glmer(sinuosity ~ ratio_accum_scaled + (1 + ratio_accum_scaled|site), mod_data_byyr, family = Gamma(link = "log"))
m6_sin <- glmer(sinuosity ~ piece_100m_scaled + (1 + piece_100m_scaled|site), mod_data_byyr, family = Gamma(link = "log"))

m7_sin <- glmer(sinuosity ~ ratio_accum_scaled + piece_100m_scaled + (1 + ratio_accum_scaled|site), mod_data_byyr, family = Gamma(link = "log"))
m8_sin <- glmer(sinuosity ~ ratio_accum_scaled * piece_100m_scaled + (1 + ratio_accum_scaled|site), mod_data_byyr, family = Gamma(link = "log"))
m9_sin <- glmer(sinuosity ~ ratio_accum_scaled + piece_100m_scaled + (1 + piece_100m_scaled|site), mod_data_byyr, family = Gamma(link = "log"))
m10_sin <- glmer(sinuosity ~ ratio_accum_scaled * piece_100m_scaled + (1 + piece_100m_scaled|site), mod_data_byyr, family = Gamma(link = "log"))

# full candidate set of models
cand.set <- list(null_sin, m1_sin, m2_sin, m3_sin, m4_sin, m5_sin, m6_sin, m7_sin, m8_sin, m9_sin, m10_sin)

```

```{r check candidate sinuosity models}
## model diagnostics with DHARMa & performance packages
# Plot simulated residuals: (QQ-plot for distribution, Residual vs. Predicted for variance)
sim_resids <- purrr::map(cand.set, ~DHARMa::simulateResiduals(fittedModel = .x))

# plotting QQ plot and predicted vs residuals
purrr::map(sim_resids, ~plot(.x, quantreg = FALSE))

# plotting just residuals
purrr::map(sim_resids, ~plot(residuals(.x)))

# recalc residuals aggregated by site
recal_sim_resids <- purrr::map(sim_resids, ~DHARMa::recalculateResiduals(.x, group = ~site))

purrr::map(recal_sim_resids, ~plot(.x, quantreg = FALSE))
purrr::map(recal_sim_resids, ~plot(residuals(.x)))

# plots various model assumptions (normality of residuals, normality of random effects, linear relationship, homogeneity of variance, multicollinearity).
purrr::map(cand.set, ~performance::check_model(.x))
```

```{r sinuosity model selection}
AICcmodavg::aictab(cand.set, modnames = c("null_sim","m1_sin","m2_sin","m3_sin", "m4_sin", "m5_sin", 
                                         "m6_sin", "m7_sin","m8_sin","m9_sin", "m10_sin"))

tab_model(m7_sin)
tab_model(m1_sin)
```

```{r plot top sinuosity model predictions}
# ## plot the null model
# # create plotting df with our sampled and estimated site means
# plot_null_sin <- mod_data_byyr %>% 
#   group_by(site) %>% 
#   summarize(sampled = mean(sinuosity)) %>% 
#   mutate(estimated = exp(coef(null_sin)$site$"(Intercept)"))
# 
# # solid lines are sampled means, dashed lines are estimated means
# ggplot(mod_data_byyr) +
#   geom_point(aes(x = ratio_accum_scaled, y = sinuosity, color = site)) +
#   geom_hline(aes(yintercept = sampled, color = site), plot_null_sin) +
#   geom_hline(aes(yintercept = estimated, color = site), plot_null_sin, linetype = "dashed") 

# plot model predictions example
plot(ggpredict(m7_sin, terms = c("ratio_accum_scaled", "piece_100m_scaled", "site"), type = "random", bias_correction = TRUE))
plot(ggpredict(m1_sin, terms = c("ratio_accum_scaled", "site"), type = "random", bias_correction = TRUE))
```


## models for complexity
```{r complexity null model}
mod_data_byyr %>% 
  ggplot(aes(x=complexity)) +
  geom_density()

mod_data_byyr %>% 
  ggplot(aes(x=sdi)) +
  geom_density()

## null model: random effect of site only reflects hierarchical design 
# this will estimate a random intercept of sinuosity for every site (i.e. mean sinuosity per site)
null_sin <- glmer(complexity ~ (1|site), mod_data_byyr, family = Gamma(link = "log"))

## model diagnostics with DHARMa & performance packages
# Plot simulated residuals: (QQ-plot for distribution, Residual vs. Predicted for variance)
plot(DHARMa::simulateResiduals(null_sin))
# plots various model assumptions (normality of residuals, normality of random effects, linear relationship, homogeneity of variance, multicollinearity).
performance::check_model(null_sin)

## model summary
summary(null_sin)
coef(null_sin)
r2(null_sin)
AICc(null_sin)

## plot the null model
# create plotting df with our sampled and estimated site means
plot_null_sin <- mod_data_byyr %>% 
  group_by(site) %>% 
  summarize(sampled = mean(sinuosity)) %>% 
  mutate(estimated = exp(coef(null_sin)$site$"(Intercept)"))

# solid lines are sampled means, dashed lines are estimated means
ggplot(mod_data_byyr) +
  geom_point(aes(x = ratio_accum_scaled, y = sinuosity, color = site)) +
  geom_hline(aes(yintercept = sampled, color = site), plot_null_sin) +
  geom_hline(aes(yintercept = estimated, color = site), plot_null_sin, linetype = "dashed") 
```



## stage 3: sensitivity

it could be good to ask whether our results change with the period of record we use (i.e. the imagery quality)
especially if we think the quality of wood data drastically improves in a given period