---
title: "wood_dynamism"
author: 'Noël Clark'
date: "2026-06-22"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      out.width = "100%")
```

```{r load libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(plotly)
library(patchwork)
library(GGally)
library(lme4)
library(DHARMa)
library(lmerTest)
library(performance)
library(MASS)
```

```{r load data, message=FALSE}
## change directory if needed ##
data_dir <- "C:/Users/nc274936/OneDrive - The University of Montana/wood_dynamism/data"

dynamism_data <- read_csv(file.path(data_dir, "Metrics/site_metrics_07022026.csv"))

dynamism_data <- dynamism_data[1:13,] %>% # for some reason we have duplicate rows in the csv
  bind_rows( # also want to add back the mod rock_WA site
    dynamism_data %>% slice(6) %>% 
      mutate(across(
        c(site_reach, site),
        ~ str_replace_all(.x, "_1", "_2")
        ))
    ) 

data_byyr <- read_csv(file.path(data_dir, "/Metrics/SelectMetricsByYear_07022026.csv"))
data_byyr <- data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach)) %>% 
  mutate(ratio_accum_to_ch_area = accum_area/channel_area)


all_data_byyr <- read.csv(file.path(data_dir, "/Metrics/ALL_metrics_07022026.csv"))
all_data_byyr <- all_data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach))
```

# Simple Plots
 
plot settings
```{r plot settings}
wood_colors = c("LW_Channel" = "#99C5B5", "no_LW_Channel" = "#BEA57D")
```

# visualizing distribution of response metrics 

support for the response values is positive & continuous
none of the response values are normally distributed - may require transformation?

support for the predictor is between 0 and 1 (proportion)
the predictor follows a negative binomial (?) distribution

```{r simple response value plots}
data_byyr %>% 
  ggplot(aes(x=sinuosity)) +
  geom_density() 

data_byyr %>% 
  ggplot(aes(x=complexity)) +
  geom_density()

data_byyr %>% 
  ggplot(aes(x=sdi)) +
  geom_density() 

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area)) +
  geom_density() 

```

## stage 1: wood as a response to site, reach, & time
```{r wood v time plots}
# how does standardized wood (the ratio of wood accumulation to channel area) change over time?
data_byyr %>%
  ggplot(aes(x=year, y=ratio_accum_to_ch_area, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach))

# how does the difference between standardized wood change within a site (between reaches) over time?
data_byyr %>%
  group_by(site, year) %>% 
  arrange_all(by_group = "reach") %>% 
  mutate(accum_ratio_diff = lag(ratio_accum_to_ch_area, 1) - ratio_accum_to_ch_area) %>% 
  dplyr::select(site, year, accum_ratio_diff) %>% 
  drop_na(accum_ratio_diff) %>% 
  ggplot(aes(x=year, y=accum_ratio_diff, color = site)) +
  geom_point() +
  geom_line(aes(group = site))
```
```{r simple wood site plots}
##### wood values by site by year #####
data_byyr %>% 
  ggplot(aes(x = year, y = ratio_accum_to_ch_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors) +
  theme(legend.position = "none") 

data_byyr %>% 
  ggplot(aes(x = year, y = wood_pieces, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

##### mean wood values by site #####
dynamism_data %>% 
  ggplot(aes(x = site, y = avg_wood_pieces_per100m, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors) +
  theme(legend.position = "none")

dynamism_data %>% 
  ggplot(aes(x = site, y = mean_wood_accum_per_area, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors)

```


## stage 2: how do sinuosity, complexity, and sdi change as a function of standardized wood?

visualizing relationship between response v predictor (ratio_accum_to_ch_area)
within a site_reach combo the relationship is linear-ish? 
```{r visualize sin comp and sdi v wood}
data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sinuosity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=complexity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sdi, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))
```

channel values by site
```{r simple channel site plots}
# channel characteristics v time by site
data_byyr %>% 
  ggplot(aes(x = year, y = channel_length, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = channel_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) + 
  scale_color_manual(values = wood_colors)
```

response values by site
```{r simple response values by site}
# response metrics v time by site
data_byyr %>% 
  ggplot(aes(x = year, y = sinuosity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = complexity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = sdi, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = net_change, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = erosion_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = deposition_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
```

dynamism values by site
```{r simple dynamism plots}
##### Sinuosity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Sinuosity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Sinuosity Dynamism vs. Avg Wood Metrics")
##### end 

##### Complexity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Complexity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Complexity Dynamism vs. Avg Wood Metrics")
##### end

##### SDI vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "SDI Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "SDI Dynamism vs. Avg Wood Metrics")
##### end

##### Net change vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Net Change Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Net Change Dynamism vs. Avg Wood Metrics")
##### end
```

# correlation plots and test 

ggpairs uses the pearson correlation test by default
```{r correlation plot test}
# ggpairs uses pearson cor test by default 
data_byyr %>%
  dplyr::select(sinuosity, complexity, sdi, channel_length, channel_area, accum_area, ratio_accum_to_ch_area, wood_pieces) %>% 
  ggpairs(progress = F)

dynamism_data %>% 
  dplyr::select(sinuosity_dynamism, complexity_dynamism, sdi_dynamism, net_change_dynamism, avg_wood_pieces_per100m, mean_wood_accum_per_area) %>% 
  ggpairs(progress = F)
```

# Models

## base model for sinuosity
"random effects should reflect the design, but complexity should be justified by the amount of information available"
```{r sinuosity base model}
# byyr_mod_data <- data_byyr %>% 
#   filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
#   mutate(site = str_replace_all(site, "_[:digit:]$", ""))

data_byyr <- data_byyr %>% 
  mutate(site = factor(site),
         reach = factor(reach))

mod_data_byyr <- data_byyr %>% 
  mutate(ratio_accum_scaled = scale(ratio_accum_to_ch_area),
         year_c = year - mean(year))

mod_data_byyr %>% 
  ggplot(aes(x=ratio_accum_scaled)) +
  geom_density()

# simplest model that reflects the hierarchical design of the study 
base_sin <- lmer(sinuosity ~ year * reach + ratio_accum_to_ch_area + (1|site), data_byyr)
base_sin_scaled <- lmer(sinuosity ~ year_c * reach + ratio_accum_scaled + (1|site), mod_data_byyr)

plot(simulateResiduals(base_sin_scaled))
summary(base_sin_scaled)
coef(base_sin_scaled)
r2(base_sin_scaled)

site_intercepts <- coef(base_sin_scaled)$site$"(Intercept)"

mod_data_byyr %>% 
  ggplot(aes(x=ratio_accum_scaled, y=sinuosity, color = site)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  geom_abline(slope = 0.2127, intercept = 2.758) +
  geom_abline(slope = (0.2127-0.2605), intercept = 2.758, color = "red") +
  geom_smooth(formula = y ~ x)

mod_data_byyr %>% 
  ggplot(aes(x=ratio_accum_scaled, y=sinuosity, color = site_reach)) +
  geom_point() +
  geom_smooth(formula = y ~ x)

```


## stage 3: sensitivity

it could be good to ask whether our results change with the period of record we use (i.e. the imagery quality)
especially if we think the quality of wood data drastically improves in a given period