---
title: "wood_dynamism"
author: 'Noël Clark'
date: "2026-06-22"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      out.width = "100%")
```

```{r load libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(plotly)
library(patchwork)
library(GGally)
library(lme4)
library(DHARMa)
library(lmerTest)
library(performance)
library(MASS)
```

```{r load data, message=FALSE}
##### change directory if needed ######
data_dir <- "C:/Users/nc274936/OneDrive - The University of Montana/wood_dynamism/data"
setwd("~/Library/CloudStorage/OneDrive-TheUniversityofMontana/Wood+/Wood+") # Kelley's working directory

dynamism_data <- read_csv(file.path(data_dir, "Metrics/site_metrics_07022026.csv"))
dynamism_data <- read_csv("site_metrics_07022026.csv") # Kelley's path

dynamism_data <- dynamism_data[1:13,] %>% # for some reason we have duplicate rows in the csv
  bind_rows( # also want to add back the mod rock_WA site
    dynamism_data %>% slice(6) %>% 
      mutate(across(
        c(site_reach, site),
        ~ str_replace_all(.x, "_1", "_2")
        ))
    ) 

data_byyr <- read_csv(file.path(data_dir, "/Metrics/SelectMetricsByYear_07022026.csv"))
data_byyr <- read_csv("SelectMetricsByYear_07022026.csv") # Kelley's path
data_byyr <- data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach)) %>% 
  mutate(ratio_accum_to_ch_area = accum_area/channel_area)


all_data_byyr <- read.csv(file.path(data_dir, "/Metrics/ALL_metrics_07022026.csv"))
all_data_byyr <- read.csv("ALL_metrics_07022026.csv") # Kelley's path
all_data_byyr <- all_data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach))
```

# Simple Plots

## plot settings
```{r plot settings}
wood_colors = c("LW_Channel" = "#99C5B5", "no_LW_Channel" = "#BEA57D")
```

## visualizing distribution of response metrics 
```{r simple response value plots}
data_byyr %>% 
  ggplot(aes(x=sinuosity)) +
  geom_density() 

data_byyr %>% 
  ggplot(aes(x=complexity)) +
  geom_density()

data_byyr %>% 
  ggplot(aes(x=sdi)) +
  geom_density() 

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area)) +
  geom_density() 

```

## stage 1: wood as a response to site, reach, & time
```{r}
# how does standardized wood (the ratio of wood accumulation to channel area) change over time?
data_byyr %>%
  ggplot(aes(x=year, y=ratio_accum_to_ch_area, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach))

```


## stage 2: how do sinuosity, complexity, and sdi change as a function of standardized wood

## visualizing relationship between response v predictor (ratio_accum_to_ch_area)
## within a site_reach combo the relationship is linear-ish? 
```{r visualize sin comp and sdi v wood}
data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sinuosity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=complexity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sdi, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))
```

## wood values by site
```{r simple wood site plots}
##### wood values by site by year #####
data_byyr %>% 
  ggplot(aes(x = year, y = ratio_accum_to_ch_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors) +
  theme(legend.position = "none") 

data_byyr %>% 
  ggplot(aes(x = year, y = wood_pieces, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

##### mean wood values by site #####
dynamism_data %>% 
  ggplot(aes(x = site, y = avg_wood_pieces_per100m, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors) +
  theme(legend.position = "none")

dynamism_data %>% 
  ggplot(aes(x = site, y = mean_wood_accum_per_area, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors)

```

## channel values by site
```{r simple channel site plots}
##### channel characteristics v time by site #####
data_byyr %>% 
  ggplot(aes(x = year, y = channel_length, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = channel_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) + 
  scale_color_manual(values = wood_colors)
```

## response values by site
```{r simple response values by site}
##### response metrics v time by site #####
data_byyr %>% 
  ggplot(aes(x = year, y = sinuosity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = complexity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = sdi, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = net_change, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = erosion_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = deposition_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
```

## dynamism values by site
```{r simple dynamism plots}
##### Sinuosity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Sinuosity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Sinuosity Dynamism vs. Avg Wood Metrics")
##### end 

##### Complexity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Complexity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Complexity Dynamism vs. Avg Wood Metrics")
##### end

##### SDI vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "SDI Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "SDI Dynamism vs. Avg Wood Metrics")
##### end

##### Net change vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Net Change Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Net Change Dynamism vs. Avg Wood Metrics")
##### end
```
```{r simple dynamism plots pt. 2}
dynamism_slay <- dynamism_data %>%
  pivot_longer(c(sinuosity_dynamism, sdi_dynamism, net_change_dynamism),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  mutate(Value_scaled = scale(Value))

ggplot(dynamism_slay, aes(avg_wood_pieces_per100m, Value_scaled, color = Metric)) +
  geom_line(size = 1)




```


## correlation plots and test 
### ggpairs uses the pearson correlation test by default
```{r correlation plot test}
# ggpairs uses pearson cor test by default 
data_byyr %>%
  mutate(wood_pieces_per_100m = (wood_pieces*100)/(channel_length),
         wood_accum_per_area = accum_area/channel_area) %>% 
  dplyr::select(sinuosity, complexity, sdi, channel_length, channel_area, wood_pieces_per_100m, wood_accum_per_area) %>% 
  ggpairs(progress = F)

dynamism_data %>% 
  dplyr::select(sinuosity_dynamism, complexity_dynamism, sdi_dynamism, net_change_dynamism, avg_wood_pieces_per100m, mean_wood_accum_per_area) %>% 
  ggpairs(progress = F)
```

## base model for sinuosity
# "random effects should reflect the design, but complexity should be justified by the amount of information available"
```{r sinuosity base model}
# byyr_mod_data <- data_byyr %>% 
#   filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
#   mutate(site = str_replace_all(site, "_[:digit:]$", ""))

data_byyr <- data_byyr %>% 
  mutate(site = factor(site),
         reach = factor(reach))

mod_data_byyr <- data_byyr %>% 
  mutate(log_sin = log(sinuosity))

mod_data_byyr %>% 
  ggplot(aes(x=log_sin)) +
  geom_density()

# simplest model that reflects the hierarchical design of the study 
base_sin <- lmer(log_sin ~ year * reach + ratio_accum_to_ch_area + (1|site), mod_data_byyr)

plot(simulateResiduals(base_sin))
summary(base_sin)
coef(base_sin)
r2(base_sin)

site_intercepts <- coef(base_sin)$site$"(Intercept)"

data_byyr %>% 
  ggplot(aes(x=site, y=sinuosity)) +
  geom_point() +
  geom_hline(yintercept = 3.85) +
  geom_hline(yintercept = site_intercepts, linetype = "dashed", color = "gray32")

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sinuosity)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  geom_hline(yintercept = 3.85) +
  geom_hline(yintercept = site_intercepts, linetype = "dashed", color = "gray32")
```


