---
title: "wood_dynamism"
author: 'Noël Clark'
date: "2026-06-22"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      out.width = "100%")
```

```{r load libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(plotly)
library(patchwork)
library(GGally)
library(lme4)
library(DHARMa)
library(lmerTest)
library(performance)
library(MASS)
library(see)
library(AICcmodavg)
```

```{r load data, message=FALSE}
## change directory if needed ##
data_dir <- "C:/Users/nc274936/OneDrive - The University of Montana/wood_dynamism/data"
setwd("~/Library/CloudStorage/OneDrive-TheUniversityofMontana/Wood+/Wood+") # Kelley's working directory

dynamism_data <- read_csv(file.path(data_dir, "Metrics/site_metrics_07022026.csv"))
dynamism_data <- read_csv("site_metrics_07022026.csv") # Kelley's path

dynamism_data <- dynamism_data[1:13,] %>% # for some reason we have duplicate rows in the csv
  bind_rows( # also want to add back the mod rock_WA site
    dynamism_data %>% slice(6) %>% 
      mutate(across(
        c(site_reach, site),
        ~ str_replace_all(.x, "_1", "_2")
        ))
    ) 

data_byyr <- read_csv(file.path(data_dir, "/Metrics/SelectMetricsByYear_07022026.csv"))
data_byyr <- read_csv("SelectMetricsByYear_07022026.csv") # Kelley's path
data_byyr <- data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach)) %>% 
  mutate(ratio_accum_to_ch_area = accum_area/channel_area,
         site = factor(site),
         reach = factor(reach))


all_data_byyr <- read.csv(file.path(data_dir, "/Metrics/ALL_metrics_07022026.csv"))
all_data_byyr <- read.csv("ALL_metrics_07022026.csv") # Kelley's path
all_data_byyr <- all_data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach))
```

# Simple Plots
 
plot settings
```{r plot settings}
wood_colors = c("LW_Channel" = "#99C5B5", "no_LW_Channel" = "#BEA57D")
```

# visualizing distribution of response metrics 

support for the response values is positive & continuous
none of the response values are normally distributed - may require transformation?

support for the predictor is between 0 and 1 (proportion)
the predictor follows a negative binomial (?) distribution

```{r simple response value plots}
data_byyr %>% 
  ggplot(aes(x=sinuosity)) +
  geom_density() 

data_byyr %>% 
  ggplot(aes(x=complexity)) +
  geom_density()

data_byyr %>% 
  ggplot(aes(x=sdi)) +
  geom_density() 

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area)) +
  geom_density() 

data_byyr %>% 
  mutate(piece_per_100m = (wood_pieces/channel_length) * 100) %>% 
  mutate(wood_abund = ratio_accum_to_ch_area + piece_per_100m)
```

## stage 1: wood as a response to site, reach, & time
```{r wood v time plots}
# how does standardized wood (the ratio of wood accumulation to channel area) change over time?
data_byyr %>%
  ggplot(aes(x=year, y=ratio_accum_to_ch_area, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach))

# how does the difference between standardized wood change within a site (between reaches) over time?
data_byyr %>%
  group_by(site, year) %>% 
  arrange_all(by_group = "reach") %>% 
  mutate(accum_ratio_diff = lag(ratio_accum_to_ch_area, 1) - ratio_accum_to_ch_area) %>% 
  dplyr::select(site, year, accum_ratio_diff) %>% 
  drop_na(accum_ratio_diff) %>% 
  ggplot(aes(x=year, y=accum_ratio_diff, color = site)) +
  geom_point() +
  geom_line(aes(group = site))
```
```{r simple wood site plots}
##### wood values by site by year #####
data_byyr %>% 
  ggplot(aes(x = year, y = ratio_accum_to_ch_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors) +
  theme(legend.position = "none") 

data_byyr %>% 
  ggplot(aes(x = year, y = wood_pieces, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

##### mean wood values by site #####
dynamism_data %>% 
  ggplot(aes(x = site, y = avg_wood_pieces_per100m, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors) +
  theme(legend.position = "none")

dynamism_data %>% 
  ggplot(aes(x = site, y = mean_wood_accum_per_area, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors)

```


## stage 2: how do sinuosity, complexity, and sdi change as a function of standardized wood?

visualizing relationship between response v predictor (ratio_accum_to_ch_area)
within a site_reach combo the relationship is linear-ish? 
```{r visualize sin comp and sdi v wood}
data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sinuosity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=complexity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sdi, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site_reach))
```

channel values by site
```{r simple channel site plots}
# channel characteristics v time by site
data_byyr %>% 
  ggplot(aes(x = year, y = channel_length, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = channel_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) + 
  scale_color_manual(values = wood_colors)
```

response values by site
```{r simple response values by site}
# response metrics v time by site
data_byyr %>% 
  ggplot(aes(x = year, y = sinuosity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = complexity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = sdi, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = net_change, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = erosion_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = deposition_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
```

dynamism values by site
```{r simple dynamism plots}
##### Sinuosity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Sinuosity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Sinuosity Dynamism vs. Avg Wood Metrics")
##### end 

##### Complexity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Complexity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Complexity Dynamism vs. Avg Wood Metrics")
##### end

##### SDI vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "SDI Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "SDI Dynamism vs. Avg Wood Metrics")
##### end

##### Net change vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Net Change Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Net Change Dynamism vs. Avg Wood Metrics")
##### end
```

```{r simple dynamism plots pt. 2}
library(grid)
library(rcartocolor)
library(dplyr)
library(patchwork)

## CLEANING UP
# Adjusting so all 3 metrics can be plotted on same graph
dynamism_cubed <- dynamism_data %>%
  pivot_longer(c(sinuosity_dynamism, sdi_dynamism, net_change_dynamism),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  mutate(Value_scaled = (scale(Value))) # calculated a separate z-score for each dynamism metric
  # will show dynamism values relative to each other

# Essentially, does wood accumulation change dynamism relative to it's normal range?

# Looking at colorblind friendly color schemes
rcartocolor::display_carto_all()
carto_pal(7, "Fall")

# Renaming labels
dynamism_cubed <- dynamism_cubed %>%
  mutate(
    Metric = recode(
      Metric,
      sinuosity_dynamism = "Sinuosity",
      sdi_dynamism = "SDI",
      net_change_dynamism = "Net Channel Change"
    ),
    reach = recode(
      reach,
      LW_Channel = "Large Wood",
      no_LW_Channel = "Control"
    ))
    
## PLOTS
# Good looking plots but lacking detail of site and reach
ggplot(dynamism_cubed, aes(avg_wood_pieces_per100m, Value_scaled, color = Metric)) +
  geom_line(size = 1)

ggplot(dynamism_cubed, aes(mean_wood_accum_per_area, Value_scaled, color = Metric)) +
  geom_line(size = 1)

# For avg_wood_pieces_per100m
avgwoodpieces <- ggplot(
  dynamism_cubed,
  aes(
    x = avg_wood_pieces_per100m,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
  facet_wrap(~reach) +
    facet_wrap(~reach, scales = "free_x") +
    labs(
    x = "Mean wood pieces (per 100m)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()


# For mean_wood_accum_per_area
meanwood <- ggplot(
  dynamism_cubed,
  aes(
    x = mean_wood_accum_per_area,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
    facet_wrap(~reach, scales = "free_x") +
    labs(
    x = "Mean wood accumulation (per area)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()

# Fixing x-axes so you can see them on their own scale since it's wacky in the above
meanwood_freex <- ggplot(
  dynamism_cubed,
  aes(
    x = mean_wood_accum_per_area,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(
    aes(group = Metric),
    method = "lm",
    se = FALSE,
    linewidth = 1.2
  ) +
  facet_wrap(~reach, scales = "free_x") +
    labs(
    x = "Mean wood accumulation (per area)",
    #y = "Standardized dynamism (z-score)", removing so it doesn't show up twice in patchwork
    color = "Dynamism Metric",
    shape = "Site"
  ) +
  theme_bw()+
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())


## FINAL PLOT
(avgwood + meanwood_freex) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Dynamism vs. Average Wood Metrics") &
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.x = unit(0.2, "cm"),
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0.1, "cm"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  ) &
  guides(
    color = guide_legend(
      title = "Dynamism Metric",
      ncol = 1
    ),
    shape = guide_legend(
      title = "Site",
      ncol = 1,
      override.aes = list(size = 3)
    )
  )
```


# correlation plots and test 

ggpairs uses the pearson correlation test by default
```{r correlation plot test}
# ggpairs uses pearson cor test by default 
data_byyr %>%
  dplyr::select(sinuosity, complexity, sdi, channel_length, channel_area, accum_area, ratio_accum_to_ch_area, wood_pieces) %>% 
  ggpairs(progress = F)

dynamism_data %>% 
  dplyr::select(sinuosity_dynamism, complexity_dynamism, sdi_dynamism, net_change_dynamism, avg_wood_pieces_per100m, mean_wood_accum_per_area) %>% 
  ggpairs(progress = F)
```

# Models

## models for sinuosity
"random effects should reflect the design, but complexity should be justified by the amount of information available"
experimental design is such that reaches are nested within sites to account for multiple observations (across years) at reach_site combinations 
```{r sinuosity base model}
# byyr_mod_data <- data_byyr %>% 
#   filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
#   mutate(site = str_replace_all(site, "_[:digit:]$", ""))

mod_data_byyr <- data_byyr %>% 
  mutate(ratio_accum_scaled = scale(ratio_accum_to_ch_area, # scale centers the values around 0 so a value is the # of SD
                                    center = TRUE, scale = TRUE), 
         year_c = scale(year, # centers years around a mean value for year
                        center = TRUE, scale = FALSE))

mod_data_byyr %>% 
  ggplot(aes(x=ratio_accum_scaled)) +
  geom_density()

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area)) +
  geom_density()

# simplest model that reflects the hierarchical design of the study
# ratio is continuous and scaled, reach is nested within site which accounts for repeated observations over time?
base_sin <- lmer(sinuosity ~ ratio_accum_scaled + (1|site) + (1|reach:site), mod_data_byyr)

summary(base_sin)

plot(simulateResiduals(base_sin))
check_model(base_sin)
summary(base_sin)
coef(base_sin)
r2(base_sin)

AICc(base_sin)
AIC(base_sin)

mod_data_byyr %>% 
  ggplot(aes(x=ratio_accum_scaled, y=sinuosity, color = site)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  geom_abline(slope = 0.2127, intercept = 2.758) +
  geom_abline(slope = (0.2127-0.2605), intercept = 2.758, color = "red") +
  geom_smooth(formula = y ~ x)


base_sin_int <- lmer(sinuosity ~ reach * ratio_accum_scaled + (1|site), mod_data_byyr)

plot(simulateResiduals(base_sin_int))
check_model(base_sin_int)
summary(base_sin_int)
coef(base_sin_int)
r2(base_sin_int)


# reach is a random effect nested within the random effect of site
base_sin_rand <- lmer(sinuosity ~ ratio_accum_scaled + (1|site) + (1|reach:site), mod_data_byyr)

plot(simulateResiduals(base_sin_rand))
check_model(base_sin_rand)
summary(base_sin_int)
coef(base_sin_int)
r2(base_sin_int)
```


## stage 3: sensitivity

it could be good to ask whether our results change with the period of record we use (i.e. the imagery quality)
especially if we think the quality of wood data drastically improves in a given period